---
title: "fixest"
authors: "Nick Huntington-Klein, Grant McDermott, and Kyle Butts"
output: html_document
---

# Regression analysis with fixest

[**fixest**](https://lrberge.github.io/fixest) (by Laurent Bergé) is a package
designed from the ground up in C++ to make running regressions fast and
incredibly easy. It provides in-built support for a variety of linear and
non-linear models, as well as regression tables and plotting methods.

Before continuing, make sure that you have installed `fixest`. You only
have to do this once (or as often as you want to update the package).

```{r , eval = F}
# Install from CRAN (recommended) 
install.packages(fixest)

# Install the development version from GitHub (advanced) 
# Requires Rtools and the remotes package 
# remotes::install_github('lrberge/fixest')

```


Once `fixest` is installed, don't forget to load it whenever you want to
use it. Unlike Stata, you have to re-load a package every time you start a new R
session.

```{r }
# Load fixest into our current R session
library(fixest)

```


## Introduction to Fixest

The [fixest](https://lrberge.github.io/fixest/index.html) package contains a highly flexible set of tools that allow you to estimate a fairly large set of standard regression models. While the package certainly doesn't cover **every** model that exists, there is a non-negligible subset of Stata users for whom every model they've ever needed to run is covered by `fixest.`

This includes regular ol' linear regression in the `feols()` function, which builds off of the Base R standard regression function `lm(),` but also includes things like instrumental variables via 2SLS, and of course support for as many fixed effects as you'd like. `fixest` isn't limited to linear regression either, covering IV and fixed-effects support for a wide range of GLM models like logit, probit, Poisson, negative binomial, and so on in `feglm()` and `fepois().`

`fixest` covers all of this while being very fast. If you felt a speed boost going from Stata's `xtreg` to `reghdfe,` get ready for another significant improvement when moving to `fixest.`

You also get a fair amount of convenience. Adjusting your standard errors to be heteroskedasticity-robust or clustered can be a pain in other R regression functions, but it is easy in `fixest` with the `vcov` option. Regression tables, coefficient and interaction-margin plots, selecting long lists of controls without having to type them all in, lagged variables, retrieving estimated fixed effects, Wald tests, and the choice of reference for categorical variables are all made easy. You even get some stuff that's rather tricky in Stata, like automatically iterating over a bunch of model specifications, basic and staggered difference-in-difference support, or Conley standard errors.

Using `fixest` for regression starts with writing a formula. While there are plenty of bells and whistles to ad d, at its core regression formulas take the form `y ~ x1 + x2 | fe1 + fe2` where y is the outcome, x1 and x2 are predictors, and fe1 and fe2 are your sets of fixed effects.

To begin, we will use a modified dataset from the CPS with some added variables for demonstration purposes. To load the data run the following:

```{r }
# Base R reads CSVs too, but we'll use data.table here
dat = data.table::fread('https://raw.githubusercontent.com/stata2r/stata2r.github.io/main/data/cps_long.csv')

```


## Models

Unike Stata, which only ever has one active dataset in memory, remember that having multiple datasets in your global environment is the norm in R. We highlight this difference to head off a very common error for new Stata R users: you need to specify **which** dataset you're using in your model calls, e.g. `feols(..., data = dat)`. We'll see lots of examples below. At the same time, note that **fixest** allows you to set various [global options](https://lrberge.github.io/fixest/reference/index.html#section-default-values), including which dataset you want to use for all of your regressions. Again, we'll see examples below.


```{r simple-model}
feols(wage ~ educ, data = dat) 
feols(wage ~ educ + age, data = dat)

# Aside 1: `data = ...` is always the first argument 
# after the model formula. So many R users would just 
# write: 
feols(wage ~ educ, dat) 

# Aside 2: You can also set your dataset globally so 
# that you don't have to reference it each time. 
setFixest_estimation(data = dat) 
feols(wage ~ educ) 
feols(wage ~ educ + age) 
# etc.

```



```{r categorical-variables}
feols(wage ~ educ + i(treat), dat) 

# Specifying a baseline:
feols(wage ~ educ + i(treat, ref = 1), dat)

```



```{r fixed-effects}
feols(wage ~ educ | countyfips, dat) 

# Aside: fixest automatically clusters SEs by the first 
# fixed effect (if there are any). We'll get to SEs 
# later, but if you just want iid errors for a fixed 
# effect model: 
feols(wage ~ educ | countyfips, dat, vcov = 'iid') 

# Add more fixed effects... 
feols(wage ~ educ | countyfips + year, 
      dat, vcov = ~countyfips + year) 
feols(wage ~ educ | countyfips^year, 
      dat) # defaults to vcov = ~countyfips^year

```



```{r instrumental-variables}
feols(wage ~ 1 | educ ~ age, dat)  
feols(wage ~ marr | educ ~ age, dat) 

# With fixed effects (IV 1st stage always comes last) 
feols(wage ~ marr | countyfips | educ ~ age, dat)

```



```{r macros-wildcards-and-shortcuts}
ctrls = c("age", "black", "hisp", "marr") 
feols(wage ~ educ + .[ctrls], dat) 

feols(wage ~ educ + ..('^x'), dat) # ^ = starts with 
feols(wage ~ educ + ..('sp$'), dat) # $ = ends with 
feols(wage ~ educ + ..('ac'), dat) 

# Many more macro options. See `?setFixest_fml` and
# `?setFixest_estimation`. Example (reminder) where 
# you set your dataset globally, so you don't have to 
# retype `data = ...` anymore. 
setFixest_estimation(data = dat) 
feols(wage ~ educ) 
feols(wage ~ educ + .[ctrls] | statefips) 
# Etc.

```



```{r nonlinear-models}
# feglm() runs all sorts of GLM models, with the same FE features as feols()!
est1 = feglm(marr ~ age + black + hisp, data = dat, family = binomial(link = 'logit'))
est2 = feglm(marr ~ age + black + hisp | statefips + year, data = dat, family = binomial(link = 'probit'))

# fepois() is there for Poisson regression
est3 = fepois(educ ~ age + black + hisp | statefips + year, data = dat)

```



In addition to the ability to estimate a difference-in-differences design using two-way fixed effects (if the design is appropriate for that - no staggered treatment, for instance), `fixest` offers several other DID-specific tools. The below examples use generic data sets, since the CPS data used in the rest of this page is not appropriate for DID.

```{r difference-in-differences, eval = F}
# did_means provides tables of means, SEs, and treatment/control and pre/post differences for 2x2 DID
did_means(outcome + control ~ treat | post)

# sunab() produces interactions of the type that allow you to estimate the Sun & Abraham model
# for staggered treatment timing, and automatically get average treatment effects for each relative period
sunab_model = feols(y ~ control + sunab(year_treated, year))
etable(sunab_model)

```


## Interactions


```{r interact-continuous-variables}
feols(wage ~ educ:age, dat) 
feols(wage ~ educ*age, dat) 

# Polynomials 
feols(wage ~ I(age^2), dat) 
feols(wage ~ poly(age, 2, raw = TRUE))

```



```{r interact-categorical-variables}
feols(wage ~ i(treat, i.hisp), dat) 

# Aside: i() is a fixest-specific shortcut that also 
# has synergies with some other fixest functions. But 
# base R interaction operators all still work, e.g. 
feols(wage ~ factor(treat)/factor(hisp), dat) 
feols(wage ~ factor(treat)*factor(hisp), dat)

```



```{r interact-categorical-with-continuous-variables}
feols(wage ~ i(treat, age), dat) 

# Aside: i() is a fixest-specific shortcut that also 
# has synergies with some other fixest functions. But 
# base R interaction operators all still work, e.g. 
feols(wage ~ factor(treat):age, dat) 
feols(wage ~ factor(treat)/age, dat) 
feols(wage ~ factor(treat)*age, dat)

```



```{r interact-fixed-effects}
# Combine fixed effects 
feols(wage ~ educ | statefips^year, dat) 

# Varying slopes (e.g. time trend for each state) 
feols(wage ~ educ | statefips[year], dat)

```


## Standard errors

While you can specify standard errors inside the original `fixest` model call (just like Stata), a unique feature of R is that you can adjust errors for an exisiting model **on the fly**. This has [several benefits](https://grantmcdermott.com/better-way-adjust-SEs), including being much more efficient since you don't have to re-estimate your whole model. We'll try to highlight examples of both approaches below.


```{r hc}
feols(wage ~ educ, dat, vcov = 'hc1') 
feols(wage ~ educ, dat, vcov = sandwich::vcovHC) 

# Note: You can also adjust the SEs of an existing model 
m = feols(wage ~ educ, dat) 
summary(m, vcov = 'hc1')

```



```{r hac, eval = F}
feols(y ~ x, dat, vcov = 'NW', panel.id = ~unit + time)
feols(y ~ x, dat, vcov = 'NW') # if panel id is already set (see below)

```



```{r clustered}
feols(wage ~ educ | countyfips, dat) # Auto clusters by FE 
# feols(wage ~ educ | countyfips, dat, vcov = ~countyfips) # ofc can be explicit too 

# Twoway clustering etc. 
feols(wage ~ educ | countyfips + year, 
      dat, vcov = ~countyfips + year) 
# feols(wage ~ educ | countyfips + year, 
#      dat, vcov = 'twoway') ## same as above 
feols(wage ~ educ | countyfips^year, 
      dat, vcov = ~countyfips^year) 

# Reminder that you can adjust the SEs of existing 
# fixest models on-the-fly. 
m = feols(wage ~ educ | countyfips + year, dat) 
m # Clustered by countyfips (default) 
summary(m, vcov = 'twoway') 
summary(m, vcov = ~countyfips^year) 
# etc.

```



```{r conley-standard-errors}
feols(wage ~ educ, dat, vcov = conley("25 mi"))

```


## Presentation


```{r regression-table}
est1 = feols(wage ~ educ + age, dat) 
etable(est1)


est2 = feols(wage ~ educ + age + black + hisp, dat) 
etable(est1, est2) 


# SEs for existing models can be adjusted on-the-fly 
etable(est1, vcov = 'hc1') 

# Report multiple SEs for the same model 
etable(est1, vcov = list('iid', 'hc1', ~id, ~countyfips)) 

# Multi-model example 
# (Two dep. vars, stepwise coefs, varying slopes, etc.) 
est_mult = feols(c(wage, age) ~ educ + csw(hisp, black) | 
                     statefips[year], 
                 dat, vcov = ~statefips^year) 
etable(est_mult)

```



```{r joint-test-of-coefficients, eval = F}
# Rename so we can use a regular expression later
data.table::setnames(dat, c('black','hisp'),c('raceeth_black','raceeth_hisp'))
est1 = feols(wage ~ educ + age + raceeth_black + raceeth_hisp + marr, dat) 
wald(est1, c('raceeth_black','raceeth_hisp'))
wald(est1, 'raceeth_')

```



```{r coefficient-plot}
# Assume we have est1 and est2 from above 
coefplot(est1) 
coefplot(list(est1, est2))

```



```{r interaction-plot}
est1 = feols(wage ~ i(hisp, age), dat) 

# Show how effect differs by group
iplot(est1)

```


## Panel

Note you don't need to specify panel.vars if you make your data a panel dataset before running the regression using the `panel` function. For example, you can use `panel(dat, ~ id + var)`.


```{r lag-variables}
feols(wage ~ educ + l(wage, 1), dat, panel.id = ~id+year)

```



```{r lead-variables}
feols(wage ~ educ + l(wage, -1), dat, panel.id = ~id+year)

```



```{r first-difference}
feols(wage ~ educ + d(wage), dat, panel.id = ~id+year)

```

